---
title: "Workflow"
author: "Asbjørn Kloppenborg"
format: html
---

This workflow quarto document is for the study regarding validation of the ARMIE sensor node.

```{r, setup}
#| echo: false
#| message: false
#| warning: false
library(here)
library(rprojroot)
library(tidyverse)
library(gt)
library(patchwork)

root <- rprojroot::find_root(rprojroot::is_rstudio_project)
knitr::opts_knit$set(root.dir = here::here())

# Save intermediate figures in a known location
knitr::opts_chunk$set(fig.path = "figures/")

# Define subfolders
folders <- c(
  "plots",
  "plots/supplementary",
  "tables",
  "tables/supplementary"
)

# Create or reset folders
for (f in folders) {
  full_path <- file.path(root, f)

  if (dir.exists(full_path)) {
    unlink(full_path, recursive = TRUE, force = TRUE)
    message("Deleted existing folder: ", full_path)
  }

  dir.create(full_path, recursive = TRUE)
  message("Created folder: ", full_path)
}
```

```{r}
#| label: read-raw-data
#| echo: false
#| message: false
#| evaluate: false
#| include: false
source(here::here("scripts", "read_data_20250521.R"))

source(here::here("scripts", "read_data_20250522.R"))

source(here::here("scripts", "read_data_20250523.R"))

source(here::here("scripts", "read_data_20250526.R"))

source(here::here("scripts", "read_data_20250528.R"))

source(here::here("scripts", "read_data_20250603.R"))

source(here::here("scripts", "read_data_20250603clean_air.R"))

# Clear the environment

rm(list = ls())
```

```{r, source functions}
source(here("scripts", "functions_main_analysis.R"))
```

```{r, load and merge cleaned data}
#| warning: false
all_data_reference <- load_and_bind_rda(c(
  "candles20250521.rda",
  "cooking20250523.rda",
  "candles_high_RH20250526.rda",
  "cooking_high_RH20250528.rda",
  "cigarette20250603.rda",
  "clean_air20250603.rda"
))

# Few observations are missing the "source" variable, we remove these
all_data_reference <- all_data_reference |>
  filter(!is.na(source)) |>
  filter(!is.na(PM1_SMPS_SPS30)) |> # remove rows with missing PM1_SMPS_SPS30 values
  select(
    timestamp_rounded_1min, timestamp_rounded_5min, source, chamber_RH, chamber_temp,
    PM1_SMPS, PM1_SMPS_SPS30, PN25_APS_valid,
    PM1_APS_valid, PM25_APS_valid,
    PM25_dusttrak,
    PN1_SMPS_SPS30, PN25_APS_valid
  ) |>
  distinct(timestamp_rounded_1min, .keep_all = TRUE) # this is implemented to remove duplicate timestamps, which are likely due to the merging of multiple datasets with different time resolutions


# Print a summary of missing values in the reference data
print(all_data_reference |> summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |>
  arrange(desc(missing_count)))


# Read low cost sensor (LCS) data
source(here::here("scripts", "read_sensor_data.R"))

# generate a timestamp rounded to the nearest minute for merging. And select variables of interest
all_data_LCS <- all_data_LCS |>
  mutate(timestamp_rounded_1min = lubridate::floor_date(timestamp, unit = "minute")) |>
  select(timestamp_rounded_1min, timestamp_rounded_5min, sensor, SPS30_PM1, SPS30_PM2.5, SHTC3Ext_T, SHTC3Ext_RH, SHTC3_T, SHTC3_RH) |>
  filter(sensor != 0) # Sensor 0 is not working properly, we remove it

# print a summary of missing values in the LCS data
print(all_data_LCS |> summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |>
  arrange(desc(missing_count)))

# Aggregate reference data to 5-minute means
all_data_reference_5min <- all_data_reference |>
  group_by(timestamp_rounded_5min, source) |>
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")

# Merge the aggregated reference data with LCS data
all_data <- all_data_reference_5min |>
  left_join(all_data_LCS, by = "timestamp_rounded_5min") |>
  filter(!is.na(sensor) & !is.na(PM25_APS_valid))


# Inspect the number of observations per sensor
all_data |>
  count(sensor) |>
  mutate(
    (hours <- n * 5 / 60)
  ) |> # hours recorded per sensor
  arrange(desc(n)) |>
  print(n = Inf)

# inspect the number of observations per sensor per exposure source:
all_data |>
  count(sensor, source) |>
  mutate(
    (hours <- n * 5 / 60)
  ) |> # hours recorded per sensor
  arrange(source, n) |>
  print(n = Inf)

# Generate a long table with information on missingness
print(
  all_data |> summarise(across(everything(), ~ sum(is.na(.)))) |> #   Summarise missing values for each column
    pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |> # Pivot to long format
    arrange(desc(missing_count)) #  Print the summary sorted by missing count
)

# save the df as .csv
write_csv(all_data, here::here("clean_data", "all_data_main_analysis.csv"))
```


```{r, labeling}
library(labelled)

# Add labels
var_label(all_data$timestamp_rounded_1min) <- "Measurement timestamp (rounded to 1 minute, UTC)"
var_label(all_data$timestamp_rounded_5min) <- "Measurement timestamp (rounded to 5 minutes, UTC)"
var_label(all_data$source) <- "Pollution source during chamber experiment"
var_label(all_data$chamber_RH) <- "Chamber relative humidity [%]"
var_label(all_data$chamber_temp) <- "Chamber temperature [°C]"
var_label(all_data$PM1_SMPS) <- "PM1 from SMPS [µg/m³]"
var_label(all_data$PM1_SMPS_SPS30) <- "PM1 from SMPS scaled to SPS30 [µg/m³]"
var_label(all_data$PM1_APS_valid) <- "PM1 from APS (valid data only) [µg/m³]"
var_label(all_data$PM25_APS_valid) <- "PM2.5 from APS (valid data only) [µg/m³]"
var_label(all_data$PM25_dusttrak) <- "PM2.5 from DustTrak [mg/m³]" # still mg/m³!
var_label(all_data$sensor) <- "Low-cost sensor ID (numeric)"
var_label(all_data$SPS30_PM1) <- "PM1 from SPS30 [µg/m³]"
var_label(all_data$SPS30_PM2.5) <- "PM2.5 from SPS30 [µg/m³]"
var_label(all_data$SHTC3Ext_T) <- "Sensor node external temperature from SHTC3 [°C]"
var_label(all_data$SHTC3Ext_RH) <- "Sensor node external relative humidity from SHTC3 [%]"
var_label(all_data$SHTC3_T) <- "Sensor node internal temperature from SHTC3 [°C]"
var_label(all_data$SHTC3_RH) <- "Sensor node internal relative humidity from SHTC3 [%]"
save(all_data, file = here::here("clean_data", "all_data_main_analysis.rda"))
```

# This section will investigate all metrics based on the raw measurements from the SPS30, without any calibration. First the SMPS data truncated at a lower limit of 0.3 um and A

```{r}
#| label: make-reference SMPS + APS and raw SPS30
#| fig-cap: "Reference PM2.5 is constructed as PM1_SMPS + PM2.5_APS_valid (µg/m³)."
all_data_ref <- make_reference_pm25(
  all_data,
  ref_col = "PM2.5_ref",
  ref_measurements = PM1_SMPS_SPS30 + PM25_APS_valid,
  reference = "SMPS + APS"
)

all_data_ref |>
  group_by(source) |>
  summarise(min = min(PM2.5_ref), max = max(PM2.5_ref), mean = mean(PM2.5_ref), median = median(PM2.5_ref), n = n())
```

```{r}
#| label: prepare SMPS + APS and raw SPS30
#| fig-cap: "Data preparation and log-transform scaffolding."
dat <- prepare_for_modeling(all_data_ref, ref_col = "PM2.5_ref")

# Quick glimpse
dplyr::glimpse(dat)
```

```{r}
#| label: loso-sensor SMPS + APS and raw SPS30
#| fig-cap: "Leave-one-sensor-out: per-sensor performance."
loso_sens_SMPS_APS_raw <- loso_by_sensor(dat, ref_col = "PM2.5_ref", method = "raw")
loso_sens_SMPS_APS_raw$per_sensor %>% arrange(desc(R))
loso_sens_SMPS_APS_raw$per_sensor
```

```{r}
#| label: loso-sensor-plots SMPS + APS and raw SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter <- plot_scatter(loso_sens_SMPS_APS_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Calibrated SPS30 vs Reference (log-log)",
  facet_factor = "holdout_sensor"
)

p_resid <- plot_residuals_vs_rh(loso_sens_SMPS_APS_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Residuals vs RH",
  facet_factor = "holdout_sensor"
) # Sensor 145 did not have the external RH sensor working

p_ba <- plot_bland_altman(loso_sens_SMPS_APS_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Bland–Altman ",
  facet_factor = "holdout_sensor"
)

p_scatter
p_resid
p_ba
```

```{r}
#| label: loso-source SMPS + APS and raw SPS30
#| fig-cap: "Leave-one-source-out: per-source performance."
loso_src_SMPS_APS_raw <- loso_by_source(dat, ref_col = "PM2.5_ref", method = "raw")
loso_src_SMPS_APS_raw$per_source %>% arrange(desc(R))
```

```{r}
#| label: loso-source-plots SMPS + APS and raw SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter_src <- plot_scatter(loso_src_SMPS_APS_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Calibrated SPS30 vs Reference",
  facet_factor = "source"
)

p_resid_src <- plot_residuals_vs_rh(loso_src_SMPS_APS_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Residuals vs RH",
  facet_factor = "source"
)

p_ba_src <- plot_bland_altman(loso_src_SMPS_APS_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Bland–Altman",
  facet_factor = "source"
)

p_scatter_src
p_resid_src
p_ba_src
```

```{r}
#| label: summary-tables SMPS + APS and raw SPS30
#| tbl-cap: "Performance summaries across LOSO schemes."
loso_sens_SMPS_APS_raw$per_sensor %>%
  select(
    holdout_sensor, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_sensor)

loso_src_SMPS_APS_raw$per_source %>%
  select(
    holdout_source, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_source)
```

# This section uses data from the SMPS and APS as reference, as these are the most accurate instruments available in the chamber. And calibrated measurements from the SPS30. The SMPS is truncated to include bins \>0.3 um. This is done to align the measurements to the claim arange for the SPS30.

```{r}
all_data_ref |>
  mutate(source_day = paste(source, as.Date(timestamp_rounded_1min))) |>
  ggplot() +
  geom_line(aes(x = timestamp_rounded_1min, y = (PM1_SMPS_SPS30 + PM25_APS_valid), color = "PM2.5")) +
  facet_wrap(~source_day, scales = "free") +
  labs(
    title = "PM2.5 APS Valid and SMPS (>0.3 um) Over Time by Source",
    x = "Time since start",
    y = "PM2.5 (µg/m³)"
  ) +
  theme_minimal()
```

```{r}
#| label: make-reference SMPS + APS and calibrated SPS30
#| fig-cap: "Reference PM2.5 is constructed as PM1_SMPS + PM2.5_APS_valid (µg/m³)."
all_data_ref <- make_reference_pm25(
  all_data,
  ref_col = "PM2.5_ref",
  ref_measurements = PM1_SMPS_SPS30 + PM25_APS_valid,
  reference = "SMPS + APS"
)


all_data_ref |>
  group_by(source) |>
  summarise(min = min(PM2.5_ref), max = max(PM2.5_ref), mean = mean(PM2.5_ref), median = median(PM2.5_ref), n = n())
```

```{r}
#| label: prepare SMPS + APS and calibrated SPS30
#| fig-cap: "Data preparation and log-transform scaffolding."

dat <- prepare_for_modeling(all_data_ref, ref_col = "PM2.5_ref")

# Quick glimpse
dplyr::glimpse(dat)
```

```{r}
#| label: loso-sensor SMPS + APS and calibrated SPS30
#| fig-cap: "Leave-one-sensor-out: per-sensor performance."
loso_sens_SMPS_APS_lm <- loso_by_sensor(dat, ref_col = "PM2.5_ref", method = "lm")
loso_sens_SMPS_APS_lm$per_sensor %>% arrange(desc(n))
```

```{r}
#| label: loso-sensor-plots SMPS + APS and calibrated SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter <- plot_scatter(loso_sens_SMPS_APS_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Calibrated SPS30 vs Reference (log-log)",
  facet_factor = "holdout_sensor"
)

p_resid <- plot_residuals_vs_rh(loso_sens_SMPS_APS_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Residuals vs RH",
  facet_factor = "holdout_sensor"
) # Sensor 145 did not have the external RH sensor working

p_ba <- plot_bland_altman(loso_sens_SMPS_APS_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Bland–Altman ",
  facet_factor = "holdout_sensor"
)

p_scatter
p_resid
p_ba
```

```{r}
#| label: loso-source SMPS + APS and calibrated SPS30
#| fig-cap: "Leave-one-source-out: per-source performance."
loso_src_SMPS_APS_lm <- loso_by_source(dat, ref_col = "PM2.5_ref", method = "lm")
loso_src_SMPS_APS_lm$per_source %>% arrange(desc(n))
```

```{r,}
#| label: loso-source-plots SMPS + APS and calibrated SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter_src <- plot_scatter(loso_src_SMPS_APS_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Calibrated SPS30 vs Reference",
  facet_factor = "source"
)

p_resid_src <- plot_residuals_vs_rh(loso_src_SMPS_APS_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Residuals vs RH",
  facet_factor = "source"
)

p_ba_src <- plot_bland_altman(loso_src_SMPS_APS_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Bland–Altman",
  facet_factor = "source"
)

p_scatter_src
p_resid_src
p_ba_src
```

```{r, summary-tables SMPS + APS and calibrated SPS30}
#| tbl-cap: "Performance summaries across LOSO schemes."
loso_sens_SMPS_APS_lm$per_sensor %>%
  select(
    holdout_sensor, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_sensor)

loso_src_SMPS_APS_lm$per_source %>%
  select(
    holdout_source, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_source)
```

# This section uses the data from the Dusttrak as reference. And raw measurements from the SPS30.

```{r, make-reference dusttrak and raw SPS30}
#| fig-cap: "Reference PM2.5 is constructed as Dusttrak (µg/m³)."
all_data_ref <- make_reference_pm25(all_data,
  ref_col = "PM2.5_ref",
  ref_measurements = PM25_dusttrak * 1000, # Dusttrak is in mg/m3, convert to ug/m3
  reference = "Dusttrak"
)

all_data_ref |>
  group_by(source) |>
  summarise(min = min(PM2.5_ref), max = max(PM2.5_ref), mean = mean(PM2.5_ref), median = median(PM2.5_ref), n = n())
```

```{r}
all_data_ref |>
  mutate(source_day = paste(source, as.Date(timestamp_rounded_1min))) |>
  ggplot() +
  geom_line(aes(x = timestamp_rounded_1min, y = (PM25_dusttrak * 1000), color = "PM2.5")) +
  facet_wrap(~source_day, scales = "free") +
  labs(
    title = "PM2.5 Dusttrak Over Time by Source",
    x = "Time since start",
    y = "PM2.5 (µg/m³)"
  ) +
  theme_minimal()
```

```{r}
#| label: prepare dusttrak and raw SPS30
#| fig-cap: "Data preparation and log-transform scaffolding."
dat <- prepare_for_modeling(all_data_ref, ref_col = "PM2.5_ref")

# Quick glimpse
dplyr::glimpse(dat)
```

```{r}
#| label: loso-sensor dusttrak and raw SPS30
#| fig-cap: "Leave-one-sensor-out: per-sensor performance."
loso_sens_dusttrak_raw <- loso_by_sensor(dat, ref_col = "PM2.5_ref", method = "raw")
loso_sens_dusttrak_raw$per_sensor %>% arrange(desc(n))
```

```{r}
#| label: loso-sensor-plots dusttrak and raw SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter <- plot_scatter(loso_sens_dusttrak_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Calibrated SPS30 vs Reference (log-log)",
  facet_factor = "holdout_sensor"
)

p_resid <- plot_residuals_vs_rh(loso_sens_dusttrak_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Residuals vs RH",
  facet_factor = "holdout_sensor"
) # Sensor 145 did not have the external RH sensor working

p_ba <- plot_bland_altman(loso_sens_dusttrak_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Bland–Altman ",
  facet_factor = "holdout_sensor"
)

p_scatter
p_resid
p_ba
```

```{r}
#| label: loso-source dusttrak and raw SPS30
#| fig-cap: "Leave-one-source-out: per-source performance."
loso_src_dusttrak_raw <- loso_by_source(dat, ref_col = "PM2.5_ref", method = "raw")
loso_src_dusttrak_raw$per_source %>% arrange(desc(n))
```

```{r}
#| label: loso-source-plots and raw SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter_src <- plot_scatter(loso_src_dusttrak_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Calibrated SPS30 vs Reference",
  facet_factor = "source"
)

p_resid_src <- plot_residuals_vs_rh(loso_src_dusttrak_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Residuals vs RH",
  facet_factor = "source"
)

p_ba_src <- plot_bland_altman(loso_src_dusttrak_raw$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Bland–Altman",
  facet_factor = "source"
)

p_scatter_src
p_resid_src
p_ba_src
```

```{r}
#| label: summary-tables dusttrak and raw SPS30
#| tbl-cap: "Performance summaries across LOSO schemes."
loso_sens_dusttrak_raw$per_sensor %>%
  select(
    holdout_sensor, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_sensor)

loso_src_dusttrak_raw$per_source %>%
  select(
    holdout_source, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_source)
```

# This section uses the data from the Dusttrak as reference. And calibrated measurements from the SPS30.

```{r}
#| label: make-reference dusttrak and calibrated SPS30
#| fig-cap: "Reference PM2.5 is constructed as Dusttrak (µg/m³)."
all_data_ref <- make_reference_pm25(all_data,
  ref_col = "PM2.5_ref",
  ref_measurements = PM25_dusttrak * 1000, # Dusttrak is in mg/m3, convert to ug/m3
  reference = "Dusttrak"
)

all_data_ref |>
  group_by(source) |>
  summarise(min = min(PM2.5_ref), max = max(PM2.5_ref), mean = mean(PM2.5_ref), median = median(PM2.5_ref), n = n())
```

```{r}
#| label: prepare dusttrak and calibrated SPS30
#| fig-cap: "Data preparation and log-transform scaffolding."
dat <- prepare_for_modeling(all_data_ref, ref_col = "PM2.5_ref")

# Quick glimpse
dplyr::glimpse(dat)
```

```{r}
#| label: loso-sensor dusttrak and calibrated SPS30
#| fig-cap: "Leave-one-sensor-out: per-sensor performance."
loso_sens_dusttrak_lm <- loso_by_sensor(dat, ref_col = "PM2.5_ref", method = "lm")
loso_sens_dusttrak_lm$per_sensor %>% arrange(desc(n))
```

```{r}
#| label: loso-sensor-plots dusttrak and calibrated SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter <- plot_scatter(loso_sens_dusttrak_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Calibrated SPS30 vs Reference (log-log)",
  facet_factor = "holdout_sensor"
)

p_resid <- plot_residuals_vs_rh(loso_sens_dusttrak_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Residuals vs RH",
  facet_factor = "holdout_sensor"
) # Sensor 145 did not have the external RH sensor working

p_ba <- plot_bland_altman(loso_sens_dusttrak_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Sensor: Bland–Altman ",
  facet_factor = "holdout_sensor"
)

p_scatter
p_resid
p_ba
```

```{r}
#| label: loso-source dusttrak and calibrated SPS30
#| fig-cap: "Leave-one-source-out: per-source performance."
loso_src_dusttrak_lm <- loso_by_source(dat, ref_col = "PM2.5_ref", method = "lm")
loso_src_dusttrak_lm$per_source %>% arrange(desc(n))
```

```{r}
#| label: loso-source-plots dusttrak and calibrated SPS30
#| fig-cap: "Pred vs ref scatter, residuals vs RH, and log-scale Bland–Altman for all LOSO predictions."
p_scatter_src <- plot_scatter(loso_src_dusttrak_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Calibrated SPS30 vs Reference",
  facet_factor = "source"
)

p_resid_src <- plot_residuals_vs_rh(loso_src_dusttrak_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Residuals vs RH",
  facet_factor = "source"
)

p_ba_src <- plot_bland_altman(loso_src_dusttrak_lm$predictions,
  ref_col = "PM2.5_ref", pred_col = "PM2.5_pred",
  title = "LOSO by Source: Bland–Altman",
  facet_factor = "source"
)

p_scatter_src
p_resid_src
p_ba_src
```

```{r}
#| label: summary-tables dusttrak and calibrated SPS30
#| tbl-cap: "Performance summaries across LOSO schemes."
loso_sens_dusttrak_lm$per_sensor %>%
  select(
    holdout_sensor, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_sensor)

loso_src_dusttrak_lm$per_source %>%
  select(
    holdout_source, n, R, R2, MAE, RMSE, cv,
    MeanError, MeanErrorPct, mean_ref, LoA_Low, LoA_High
  ) %>%
  arrange(holdout_source)
```

#Generate a (new) table 1 with range, median, mean and SD for PM2.5 and PN2.5 concentration from SMPS+APS and PM2.5 for Dusttrak
```{r}
table1_data <- all_data |>
  select(
    timestamp_rounded_5min, source,
    PM25_dusttrak,
    PM1_SMPS_SPS30, PM25_APS_valid,
    PN1_SMPS_SPS30, PN25_APS_valid
  ) |>
  mutate(
    SMPS_APS_PN = PN1_SMPS_SPS30 + PN25_APS_valid, # /cm3
    SMPS_APS_PM = PM1_SMPS_SPS30 + PM25_APS_valid,
    Dusttrak = PM25_dusttrak * 1000
  ) |>
  filter(Dusttrak >= 5 & SMPS_APS_PM >= 5) |> # only include values above 5 ug/m3
  pivot_longer(
    cols = c(Dusttrak, SMPS_APS_PN, SMPS_APS_PM),
    names_to = "instrument",
    values_to = "Concentration"
  ) |>
  mutate(
    instrument = fct_recode(
      instrument,
      "SMPS + APS (#/cm³)" = "SMPS_APS_PN",
      "SMPS + APS (µg/m³)" = "SMPS_APS_PM",
      "DustTrak  (µg/m³)" = "Dusttrak"
    ),
    instrument = fct_relevel(
      instrument,
      "SMPS + APS (#/cm³)",
      "SMPS + APS (µg/m³)",
      "DustTrak  (µg/m³)"
    ),
    source = fct_recode(
      source,
      "Candles" = "candles",
      "Candles High RH\n(70 - 80%)" = "candles_high_RH",
      "Cigarette Smoke" = "cigarette",
      "Clean Air" = "clean_air",
      "Cooking" = "cooking",
      "Cooking High RH\n(70 - 80%)" = "cooking_high_RH"
    )
  ) |>
  group_by(source, instrument) |>
  summarise(
    mean = mean(Concentration, na.rm = TRUE),
    sd = sd(Concentration, na.rm = TRUE),
    median = median(Concentration, na.rm = TRUE),
    min = min(Concentration, na.rm = TRUE),
    max = max(Concentration, na.rm = TRUE),
    .groups = "drop"
  ) |> mutate(across(where(is.numeric),   ~ round(., 2)))
```

```{r}
# Generate table 1
table1_gt <- table1_data |>
  select(
    source, instrument, median, mean, sd, min, max
  ) |>
  gt(groupname_col = "source") |>
  tab_header(
    title = "Table 1: Summary statistics of PM and PN measurements by instrument and source"
  ) |>
  cols_label(
    source = "Source",
    instrument = "Instrument",
    median = "Median",
    mean = "Mean",
    sd = "SD",
    min = "Min",
    max = "Max"
  ) |>
  fmt_number(
    columns = c(median, mean, sd, min, max),
    decimals = 0
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  )

table1_gt

# Save table as word document
gtsave(table1_gt, "tables/summary_table1.docx", overwrite = TRUE)
```

# Prepare data before plotting figure 1

```{r, prepare figure 1}
figure1_long <- all_data |>
  select(timestamp_rounded_5min, source, sensor, SPS30_PM2.5, PM25_dusttrak, PM1_SMPS_SPS30, PM25_APS_valid) %>%
  mutate(
    SMPS_APS = PM1_SMPS_SPS30 + PM25_APS_valid,
    Dusttrak = PM25_dusttrak * 1000
  ) %>%
  pivot_longer(
    cols = c(Dusttrak, SMPS_APS),
    names_to = "instrument",
    values_to = "PM25"
  ) %>%
  mutate(
    instrument = case_when(
      instrument == "SPS30_PM2.5" ~ paste("Sensor", sensor),
      TRUE ~ instrument
    ),
    instrument = factor(instrument),
    instrument = fct_recode(
      instrument,
      "SMPS + APS" = "SMPS_APS",
      "DustTrak"   = "Dusttrak"
    ),
    instrument = fct_relevel(
      instrument,
      "SMPS + APS",
      "DustTrak"
    ),
    source = factor(source),
    source = fct_recode(
      source,
      "Candles" = "candles",
      "Candles High RH\n(70 - 80%)" = "candles_high_RH",
      "Cigarette Smoke" = "cigarette",
      "Clean Air" = "clean_air",
      "Cooking" = "cooking",
      "Cooking High RH\n(70 - 80%)" = "cooking_high_RH"
    )
  )
```

# Plot figure 1

```{r, figure 1}
#| label: figure1
#| warning: false
#| # Make sure folder exists
if (!dir.exists("plots")) dir.create("plots")
# Make sure folder exists
if (!dir.exists("plots/supplementary")) dir.create("plots/supplementary")

figure1_long_elapsed <- figure1_long %>%
  group_by(source) %>%
  arrange(timestamp_rounded_5min, .by_group = TRUE) %>%
  mutate(
    elapsed_min = as.numeric(difftime(timestamp_rounded_5min,
      first(timestamp_rounded_5min),
      units = "mins"
    ))
  ) %>%
  ungroup()

figure1 <- ggplot(
  figure1_long_elapsed,
  aes(
    x = elapsed_min, y = PM25, group = instrument, color = instrument,
    alpha = instrument, linewidth = instrument
  )
) +
  geom_line() +
  # ylim(0, 750) +
  facet_wrap(~source, scales = "free") +
  theme_minimal() +
  labs(
    # title = "PM2.5 concentrations over time (elapsed minutes per source)",
    x = "Elapsed time (minutes)",
    y = "PM2.5 (µg/m³)",
    color = "Instrument"
  ) +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  )) +
  scale_alpha_manual(values = c(
    "DustTrak" = 1,
    "SMPS + APS" = 1
  )) +
  scale_linewidth_manual(values = c(
    "DustTrak" = 1,
    "SMPS + APS" = 1
  )) +
  guides(alpha = "none", linewidth = "none")


# Print figure 1
figure1

# save figure 1
ggsave("plots/figure1.png", plot = figure1, width = 8, height = 6, dpi = 600, bg = "white")
```


#Prepare data before plotting using data from the LOSO by sensor analyses
```{r, prepare data with loso by sensor}
loso_sens_SMPS_APS_raw_figure2 <- loso_sens_SMPS_APS_raw$predictions |>
  mutate(
    method = "Uncalibrated",
    reference = "SMPS + APS"
  )

loso_sens_SMPS_APS_lm_figure2 <- loso_sens_SMPS_APS_lm$predictions |>
  mutate(
    method = "Calibrated",
    reference = "SMPS + APS"
  )

loso_sens_dusttrak_raw_figure2 <- loso_sens_dusttrak_raw$predictions |>
  mutate(
    method = "Uncalibrated",
    reference = "Dusttrak"
  )

loso_sens_dusttrak_lm_figure2 <- loso_sens_dusttrak_lm$predictions |>
  mutate(
    method = "Calibrated",
    reference = "Dusttrak"
  )

by_sensor_data <- bind_rows(
  loso_sens_SMPS_APS_raw_figure2,
  loso_sens_SMPS_APS_lm_figure2,
  loso_sens_dusttrak_raw_figure2,
  loso_sens_dusttrak_lm_figure2
) |> mutate(
  error = PM2.5_pred - PM2.5_ref,
  relative_error = error / PM2.5_ref * 100,
  source = factor(source),
  source = fct_recode(
    source,
    "Candles" = "candles",
    "Candles\nHigh RH\n(70 - 80%)" = "candles_high_RH",
    "Cigarette\nSmoke" = "cigarette",
    "Clean Air" = "clean_air",
    "Cooking" = "cooking",
    "Cooking\nHigh RH\n(70 - 80%)" = "cooking_high_RH"
  ),
  reference = factor(reference),
  reference = fct_recode(
    reference,
    "SMPS + APS" = "SMPS + APS",
    "DustTrak" = "Dusttrak"
  )
)
```

# Scatter plots with SPS30 PM2.5 readings and comparision instruments (For supplementary)

```{r, figure 2 fixed axis with loso by sensor}
#| warning: false

p_uncal_AE <- by_sensor_data |>
  filter(method == "Uncalibrated") |>
  ggplot(aes(x = PM2.5_ref, y = PM2.5_pred, color = reference)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  ylim(0, 350) +
  xlim(0, 850) +
  facet_wrap(~source) +
  labs(
    title = "SPS30 vs Reference PM2.5 Concentrations",
    subtitle = "Uncalibrated values from the SPS30 sensor",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 (µg/m³)",
    color = "Comparison"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

p_cal_AE <- by_sensor_data |>
  filter(method == "Calibrated") |>
  ggplot(aes(x = PM2.5_ref, y = PM2.5_pred, color = reference)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  ylim(0, 350) +
  xlim(0, 850) +
  facet_wrap(~source) +
  labs(
    title = "SPS30 vs Reference PM2.5 Concentrations",
    subtitle = "Calibrated values from the SPS30 sensor",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 (µg/m³)",
    color = "Comparison"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

p_scatter <- (p_uncal_AE | p_cal_AE) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_scatter
# Save them
ggsave("plots/supplementary/scatter.png", plot = p_scatter, width = 8, height = 6, dpi = 600, bg = "white")
```


plot figure RE_scatter with relative error between SPS30 and comparison instrument (For supplementary)

```{r, figure x with loso by sensor}
#| warning: false

p_uncal_RE <- by_sensor_data |>
  filter(method == "Uncalibrated") |>
  ggplot(aes(x = PM2.5_ref, y = relative_error, color = reference)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, fullrange = TRUE) +
  # geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  ylim(0, 400) +
  xlim(0, 300) +
  facet_wrap(~source) +
  labs(
    title = "Relative error of SPS30 vs Reference PM2.5",
    subtitle = "Uncalibrated values from the SPS30 sensor",
    x = "Reference PM2.5 (µg/m³)",
    y = "Relative error (%)",
    color = "Comparison"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

p_cal_RE <- by_sensor_data |>
  filter(method == "Calibrated") |>
  ggplot(aes(x = PM2.5_ref, y = relative_error, color = reference)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, fullrange = TRUE) +
  # geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  ylim(0, 400) +
  xlim(0, 300) +
  facet_wrap(~source) +
  labs(
    title = "Relative error of SPS30 vs Reference PM2.5",
    subtitle = "Calibrated values from the SPS30 sensor",
    x = "Reference PM2.5 (µg/m³)",
    y = "Relative error (%)",
    color = "Comparison"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

p_scatter_RE <- (p_uncal_RE | p_cal_RE) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_scatter_RE

# Save them
ggsave("plots/supplementary/RE_scatter.png", plot = p_scatter_RE, width = 8, height = 6, dpi = 600, bg = "white")

by_sensor_data %>%
  filter(!is.finite(relative_error))
```



```{r, table sumarizing mean and sd from Bland Altman plot with loso by sensor}
#| warning: false

table2_data_clean <- by_sensor_data |>
  filter(PM2.5_ref >= 5) |>
  mutate(
    PM2.5_ref = ifelse(PM2.5_ref == 0, 0.001, PM2.5_ref) # avoid division by zero
  ) |>
  group_by(reference, method) |>
  summarise(
    # core error terms
    mae = mean(abs(error), na.rm = TRUE),
    rmse = sqrt(mean(error^2, na.rm = TRUE)),
    mean_error = mean(error, na.rm = TRUE),
    sd_error = sd(error, na.rm = TRUE),

    # relative measures
    mean_ref = mean(PM2.5_ref, na.rm = TRUE),
    mean_sps30 = mean(PM2.5_pred, na.rm = TRUE),
    err_pct = mean(error / PM2.5_ref * 100, na.rm = TRUE),

    # correlation and variation
    r = cor(PM2.5_pred, PM2.5_ref, use = "complete.obs"),
    r2 = r^2,
    cv = sd(error, na.rm = TRUE) / mean(error, na.rm = TRUE),

    # limits of agreement
    loa_low = mean_error - 1.96 * sd_error,
    loa_high = mean_error + 1.96 * sd_error,
    .groups = "drop"
  )

# Build grouped GT table
table_2 <- table2_data_clean |>
  select(reference, method, r, err_pct, mean_error, loa_low, loa_high) |>
  gt(groupname_col = "Reference") |>
  tab_header(
    title = "Table 1: Performance metrics based on the leave-one-sensor-out approach",
    # subtitle = "Against reference instruments (SMPS + APS, DustTrak), stratified by source type and calibration method"
  ) |>
  cols_label(
    reference = "Reference",
    method = "Method",
    r = "Pearson's r",
    err_pct = "Mean Relative Error (%)",
    mean_error = "Mean Error (µg/m³)",
    loa_low = "LoA Low (µg/m³)",
    loa_high = "LoA High (µg/m³)" 
    )|> 
   fmt_number(
    columns = c(r, err_pct, mean_error, loa_low, loa_high),
    decimals = 2, # choose decimals you want
    use_seps = TRUE, # adds thousand separators
    drop_trailing_zeros = TRUE
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  )

table_2

# Save table as word document
gtsave(table_2, "tables/table2.docx", overwrite = TRUE)
```


# Bland-altman plots for main text. (Figure 2)
```{r, Bland altman plot with per-facet stats with loso by sensor without source specific LoA by sensor data}
bland_altman_data <- by_sensor_data |>
  filter(PM2.5_ref >= 5) |>
  mutate(sensor = factor(sensor)) |>
  group_by(reference, method) |>
  mutate(
    ref_val = PM2.5_ref,
    diff_val = error,
    mean_diff = mean(diff_val, na.rm = TRUE),
    sd_diff = sd(diff_val, na.rm = TRUE),
    holdout_sensor = factor(holdout_sensor)
  )

BA_plot_uncal <- bland_altman_data |>
  filter(method == "Uncalibrated") |>
  ggplot(aes(x = ref_val, y = diff_val, color = source)) +
  geom_point(alpha = 0.4) +
  geom_smooth(aes(group = 1), method = "lm", color = "black", se = TRUE, fullrange = TRUE) +
  geom_hline(aes(yintercept = mean_diff), color = "#E69F00") +
  geom_hline(aes(yintercept = mean_diff + 1.96 * sd_diff), linetype = 3) +
  geom_hline(aes(yintercept = mean_diff - 1.96 * sd_diff), linetype = 3) +
  geom_hline(yintercept = 0, linetype = 2) +
  xlim(0, 400) +
  ylim(-200, 200) +
  facet_wrap(~reference) +
  labs(
    # title = "Bland–Altman Plot of SPS30 vs Reference PM2.5 Concentrations",
    subtitle = "Uncalibrated",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)"
  ) +
  theme_minimal()


BA_plot_cal <- bland_altman_data |>
  filter(method == "Calibrated") |>
  ggplot(aes(x = ref_val, y = diff_val, color = source)) +
  geom_point(alpha = 0.4) +
  geom_smooth(aes(group = 1), method = "lm", color = "black", se = TRUE, fullrange = TRUE) +
  geom_hline(aes(yintercept = mean_diff), color = "#E69F00") +
  geom_hline(aes(yintercept = mean_diff + 1.96 * sd_diff), linetype = 3, color = "#E69F00") +
  geom_hline(aes(yintercept = mean_diff - 1.96 * sd_diff), linetype = 3, color = "#E69F00") +
  geom_hline(yintercept = 0, linetype = 2) +
  xlim(0, 400) +
  ylim(-200, 200) +
  facet_wrap(~reference) +
  labs(
    # title = "Bland–Altman Plot of SPS30 vs Reference PM2.5 Concentrations",
    subtitle = "Calibrated",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)"
  ) +
  theme_minimal()

# Cowplot: extract a single legend and glue it under the combined panels
library(cowplot)

# 1. Extract the legend from one of the plots (they share the same scale)
legend <- cowplot::get_legend(
  BA_plot_uncal + theme(legend.position = "bottom")
)

# 2. Combine the two plots without legends
row <- cowplot::plot_grid(
  BA_plot_uncal + theme(legend.position = "none"),
  BA_plot_cal + theme(legend.position = "none"),
  ncol = 2, align = "hv"
)

# 3. Add the shared legend underneath
BA_plot <- cowplot::plot_grid(
  row,
  legend,
  ncol = 1,
  rel_heights = c(1, 0.1)  # adjust legend size
)

BA_plot

ggsave("plots/blandaltman_figure2.png", plot = BA_plot, width = 10, height = 6, dpi = 600, bg = "white")
```
# Bland-altman plots for supplementary.
These plots are faceted by source and method, with LoA and mean error calculated per source using the by_sensor data

```{r, Bland altman plot with per-facet stats with loso by sensor with source specific LoA}
bland_altman_data_spec_loa <- by_sensor_data %>%
  mutate(
    ref_val = PM2.5_ref,
    diff_val = PM2.5_pred - PM2.5_ref
  )

summary_stats <- bland_altman_data_spec_loa %>%
  group_by(source, reference, method) %>%
  summarise(
    mean_diff = mean(diff_val, na.rm = TRUE),
    sd_diff = sd(diff_val, na.rm = TRUE),
    .groups = "drop"
  )

bland_altman_plot_spec_loa_uncal <- bland_altman_data_spec_loa %>%
  left_join(summary_stats, by = c("source", "reference", "method")) %>%
  filter(method == "Uncalibrated") %>%
  ggplot(aes(x = ref_val, y = diff_val, color = reference)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_hline(aes(yintercept = mean_diff, color = reference), linetype = 1) +
  geom_hline(aes(yintercept = mean_diff + 1.96 * sd_diff, color = reference), linetype = 3) +
  geom_hline(aes(yintercept = mean_diff - 1.96 * sd_diff, color = reference), linetype = 3) +
  geom_hline(yintercept = 0, linetype = 2) +
  xlim(0, 400) +
  ylim(-200, 200) +
  facet_wrap(source ~ method) +
  labs(
    #   title = "Bland–Altman Plot of SPS30 vs Reference PM2.5 Concentrations",
    #   subtitle = "Calibrated SPS30 measurements (Leave one sensor out)",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

bland_altman_plot_spec_loa_cal <- bland_altman_data_spec_loa %>%
  left_join(summary_stats, by = c("source", "reference", "method")) %>%
  filter(method == "Calibrated") %>%
  ggplot(aes(x = ref_val, y = diff_val, color = reference)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_hline(aes(yintercept = mean_diff, color = reference), linetype = 1) +
  geom_hline(aes(yintercept = mean_diff + 1.96 * sd_diff, color = reference), linetype = 3) +
  geom_hline(aes(yintercept = mean_diff - 1.96 * sd_diff, color = reference), linetype = 3) +
  geom_hline(yintercept = 0, linetype = 2) +
  xlim(0, 400) +
  ylim(-200, 200) +
  facet_wrap(source ~ method) +
  labs(
    # title = "Bland–Altman Plot of SPS30 vs Reference PM2.5 Concentrations",
    # subtitle = "Calibrated SPS30 measurements (Leave one sensor out)",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

BA_plot_source_by_sensor <- (bland_altman_plot_spec_loa_uncal | bland_altman_plot_spec_loa_cal) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  labs(
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)"
  )


BA_plot_source_by_sensor
# Save them in supplementary
ggsave("plots/supplementary/bland_altman_source_specific_Loa_by_sensor.png", plot = BA_plot_source_by_sensor, width = 8, height = 6, dpi = 600, bg = "white")
```

#Plot boxplot with relative errors (Figure 3) not including clean air and non-finite relative errors

```{r, boxplot with relative errors}
#| warning: false

# Create a variable splitting PM2.5_ref into two groups
boxplot_data <- by_sensor_data |>
  filter(source != "Clean Air") |>
  mutate(
    PM25_group = if_else(PM2.5_pred <= 5, "PM2.5 =< 5 µg/m³", "PM2.5 > 5 µg/m³")
  )

# Combined boxplot for uncalibrated and calibrated
p_box_RE <- boxplot_data |>
  ggplot(aes(x = source, y = relative_error, fill = PM25_group)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(reference ~ method) +
  scale_fill_manual(
    values = c("PM2.5 =< 5 µg/m³" = "#009E73", "PM2.5 > 5 µg/m³" = "#CC79A7")
  ) +
  coord_cartesian(ylim = c(-200, 600)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, vjust = 1),
    legend.position = "bottom",
    legend.box = "horizontal"
  ) +
  labs(
    # title = "Distribution of Relative Error by Source and PM2.5 Concentration Range",
    # subtitle = "Comparison of uncalibrated and calibrated SPS30 measurements",
    x = "Source type",
    y = "Relative error (%)",
    fill = "SPS30 PM2.5"
  )

p_box_RE
# save the combined boxplot
ggsave("plots/figure3_RE_boxplot_combined.png", plot = p_box_RE, width = 10, height = 8, dpi = 600, bg = "white")
```
# Boxplot with relative errors including clean air for supplementary

```{r, boxplot with relative errors including clean air for supplementary}
# Create a variable splitting PM2.5_ref into two groups
boxplot_data <- by_sensor_data |>
  mutate(
    PM25_group = if_else(PM2.5_pred <= 5, "PM2.5 =< 5 µg/m³", "PM2.5 > 5 µg/m³")
  )

# Combined boxplot for uncalibrated and calibrated
p_box_RE_sup <- boxplot_data |>
  ggplot(aes(x = source, y = relative_error, fill = PM25_group)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(reference ~ method) +
  scale_fill_manual(
    values = c("PM2.5 =< 5 µg/m³" = "#009E73", "PM2.5 > 5 µg/m³" = "#CC79A7")
  ) +
  coord_cartesian(ylim = c(-200, 600)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, vjust = 1),
    legend.position = "bottom",
    legend.box = "horizontal"
  ) +
  labs(
    title = "Distribution of Relative Error by Source and PM2.5 Concentration Range",
    subtitle = "Comparison of uncalibrated and calibrated SPS30 measurements",
    x = "Source type",
    y = "Relative error (%)",
    fill = "SPS30 PM2.5"
  )

p_box_RE_sup
# save the combined boxplot
ggsave("plots/supplementary/RE_boxplot_supplementary.png", plot = p_box_RE_sup, width = 10, height = 8, dpi = 600, bg = "white")
```


# Supplementary table describing those points of relative error not included in the Box plot 

```{r}
#| warning: false

table_2S_data <- by_sensor_data %>%
  filter(source == "Clean Air" | !is.finite(relative_error)) %>%
  arrange(source, reference, method, PM2.5_ref, PM2.5_pred) %>%
  select(source, reference, method, sensor, timestamp_rounded_1min, PM2.5_ref, PM2.5_pred, relative_error)

df <- table_2S_data |> count(source, reference, method)


table_2S <- df |>
  arrange(source, reference, method) |>
  gt(groupname_col = "source") |>
  tab_header(
    title = "Number of Observations Excluded or in 'Clean Air' Category",
    subtitle = "Grouped by comparison instrument and calibration method"
  ) |>
  cols_label(
    reference = "Comparison",
    method    = "Method",
    n         = "Number of Observations"
  ) |>
  fmt_number(
    columns = n,
    decimals = 0,
    use_seps = TRUE
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  )

table_2S

gtsave(table_2S, "tables/supplementary/table_2S.docx")
```

# Prepare data before plotting using data from the LOSO by source analyses
```{r, prepare data with loso by source}
loso_src_SMPS_APS_raw_figure2 <- loso_src_SMPS_APS_raw$predictions |>
  mutate(
    method = "Uncalibrated",
    reference = "SMPS + APS"
  )

loso_src_SMPS_APS_lm_figure2 <- loso_src_SMPS_APS_lm$predictions |>
  mutate(
    method = "Calibrated",
    reference = "SMPS + APS"
  )

loso_src_dusttrak_raw_figure2 <- loso_src_dusttrak_raw$predictions |>
  mutate(
    method = "Uncalibrated",
    reference = "Dusttrak"
  )

loso_src_dusttrak_lm_figure2 <- loso_src_dusttrak_lm$predictions |>
  mutate(
    method = "Calibrated",
    reference = "Dusttrak"
  )

by_source_data <- bind_rows(
  loso_src_SMPS_APS_raw_figure2,
  loso_src_SMPS_APS_lm_figure2,
  loso_src_dusttrak_raw_figure2,
  loso_src_dusttrak_lm_figure2
) |> mutate(
  error = PM2.5_pred - PM2.5_ref,
  relative_error = error / PM2.5_ref * 100,
  source = factor(source),
  source = fct_recode(
    source,
    "Candles" = "candles",
    "Candles\nHigh RH\n(70 - 80%)" = "candles_high_RH",
    "Cigarette\nSmoke" = "cigarette",
    "Clean Air" = "clean_air",
    "Cooking" = "cooking",
    "Cooking\nHigh RH\n(70 - 80%)" = "cooking_high_RH"
  ),
  reference = factor(reference),
  reference = fct_recode(
    reference,
    "SMPS + APS" = "SMPS + APS",
    "DustTrak" = "Dusttrak"
  )
)
```
 
# Generate Table 2 summarizing mean and sd from Bland Altman plot with loso by source data

```{r, Table 2}
table3_data <- by_source_data |>
  filter(PM2.5_ref >= 5) |>
  mutate(
    PM2.5_ref = ifelse(PM2.5_ref == 0, 0.001, PM2.5_ref) # avoid division by zero
  ) |>
  group_by(reference, method, source) |>
  summarise(
    # core error terms
    mae = mean(abs(error), na.rm = TRUE),
    rmse = sqrt(mean(error^2, na.rm = TRUE)),
    mean_error = mean(error, na.rm = TRUE),
    sd_error = sd(error, na.rm = TRUE),

    # relative measures
    mean_ref = mean(PM2.5_ref, na.rm = TRUE),
    mean_sps30 = mean(PM2.5_pred, na.rm = TRUE),
    err_pct = mean(error / PM2.5_ref * 100, na.rm = TRUE),

    # correlation and variation
    r = cor(PM2.5_pred, PM2.5_ref, use = "complete.obs"),
    r2 = r^2,
    cv = sd(error, na.rm = TRUE) / mean(error, na.rm = TRUE),

    # limits of agreement
    loa_low = mean_error - 1.96 * sd_error,
    loa_high = mean_error + 1.96 * sd_error,
    .groups = "drop"
  )


# Build grouped GT table
table_3 <- table3_data |>
  select(source, reference, method, r, err_pct, mean_error, loa_low, loa_high) |>
  arrange(source, reference, method) |>
  gt(groupname_col = "source") |>
  tab_header(title = "Table 3: Performance metrics based on the leave-one-source-out approach") |>
  cols_label(
    reference  = "Comparison",
    method     = "Method",
    r          = "Pearson's r",
    err_pct    = "Mean Relative Error (%)",
    mean_error = "Mean Error (µg/m³)",
    loa_low    = "LoA Low (µg/m³)",
    loa_high   = "LoA High (µg/m³)"
  ) |>
  # format all numeric columns nicely
  fmt_number(
    columns = c(r, err_pct, mean_error, loa_low, loa_high),
    decimals = 2, # choose decimals you want
    use_seps = TRUE, # adds thousand separators
    drop_trailing_zeros = TRUE
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  )

table_3

# Save table as word document
gtsave(table_3, "tables/table3.docx", overwrite = TRUE)
```

#Bland-altman plots (Figure4) for main text using loso by source data with source specific LoA and mean error

```{r, bland-Altman based on by source data}
#| warning: false

# Make Bland–Altman data match the summarised table
bland_altman_data_spec_loa_by_source <- by_source_data %>%
  filter(PM2.5_ref >= 5) %>% # same cutoff as in table
  mutate(
    PM2.5_ref = ifelse(PM2.5_ref == 0, 0.001, PM2.5_ref), # same zero handling
    ref_val = PM2.5_ref
  )

# Summarise to match table metrics exactly
summary_stats <- bland_altman_data_spec_loa_by_source %>%
  group_by(source, reference, method) %>%
  summarise(
    mean_diff = mean(error, na.rm = TRUE),
    sd_diff = sd(error, na.rm = TRUE),
    loa_low = mean_diff - 1.96 * sd_diff,
    loa_high = mean_diff + 1.96 * sd_diff,
    .groups = "drop"
  )

# Check that this summary now aligns 1:1 with the GT table
# (each combination of source x reference x method should appear once)
print(summary_stats)

bland_altman_data_spec_loa_by_source_uncal <- bland_altman_data_spec_loa_by_source %>%
  left_join(summary_stats, by = c("source", "reference", "method")) %>%
  filter(method == "Uncalibrated") %>%
  ggplot(aes(x = ref_val, y = error, color = reference)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  geom_hline(aes(yintercept = mean_diff, color = reference), linetype = 1) +
  geom_hline(aes(yintercept = loa_high, color = reference), linetype = 3) +
  geom_hline(aes(yintercept = loa_low, color = reference), linetype = 3) +
  geom_hline(yintercept = 0, linetype = 2) +
  xlim(0, 400) +
  ylim(-200, 200) +
  facet_wrap(~source) +
  labs(
    subtitle = "Uncalibrated",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)",
    color = "Comparison"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

bland_altman_data_spec_loa_by_source_cal <- bland_altman_data_spec_loa_by_source %>%
  left_join(summary_stats, by = c("source", "reference", "method")) %>%
  filter(method == "Calibrated") %>%
  ggplot(aes(x = ref_val, y = error, color = reference)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  geom_hline(aes(yintercept = mean_diff, color = reference), linetype = 1) +
  geom_hline(aes(yintercept = loa_high, color = reference), linetype = 3) +
  geom_hline(aes(yintercept = loa_low, color = reference), linetype = 3) +
  geom_hline(yintercept = 0, linetype = 2) +
  xlim(0, 400) +
  ylim(-200, 200) +
  facet_wrap(~source) +
  labs(
    subtitle = "Calibrated",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)",
    color = "Comparison"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "DustTrak" = "#E69F00",
    "SMPS + APS" = "#56B4E9"
  ))

BA_plot_source <- (
  bland_altman_data_spec_loa_by_source_uncal | bland_altman_data_spec_loa_by_source_cal
) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  labs(
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 − Reference PM2.5 (µg/m³)"
  )


BA_plot_source

ggsave("plots/figure4_bland_altman_source_specific_Loa_by_source.png", plot = BA_plot_source, width = 8, height = 6, dpi = 600, bg = "white")
```


### Supplementary


```{r, investigate the fan observed in the Bland-Altman plot with loso by sensor}
#| warning: false

bland_altman_data |>
  filter(reference == "DustTrak" & method == "Uncalibrated") |>
  ggplot(aes(x = ref_val, y = abs(diff_val))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "#E69F00") +
  labs(
    title = "Data from Bland-Altman Plot: Uncalibrated SPS30 vs DustTrak",
    x = "Reference PM2.5 (µg/m³)", y = "|Difference| (µg/m³)"
  )

bland_altman_data |>
  filter(reference == "DustTrak" & method == "Calibrated") |>
  ggplot(aes(x = ref_val, y = abs(diff_val))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", color = "#E69F00") +
  labs(
    title = "Data from Bland-Altman Plot: Calibrated SPS30 vs DustTrak",
    x = "Reference PM2.5 (µg/m³)", y = "|Difference| (µg/m³)"
  )

bland_altman_data |>
  filter(reference == "SMPS + APS" & method == "Uncalibrated") |>
  ggplot(aes(x = ref_val, y = abs(diff_val))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", color = "#56B4E9") +
  labs(
    title = "Data from Bland-Altman Plot: Uncalibrated SPS30 vs SMPS + APS",
    x = "Reference PM2.5 (µg/m³)", y = "|Difference| (µg/m³)"
  )

bland_altman_data |>
  filter(reference == "SMPS + APS" & method == "Calibrated") |>
  ggplot(aes(x = ref_val, y = abs(diff_val))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", color = "#56B4E9") +
  labs(
    title = "Data from Bland-Altman Plot: Calibrated SPS30 vs SMPS + APS",
    x = "Reference PM2.5 (µg/m³)", y = "|Difference| (µg/m³)"
  )
```



```{r, investigating the effect of calibration on clean air figure 2 with loso by sensor}
#| warning: false

p_clean_air <- by_sensor_data |>
  filter(source == "Clean Air") |>
  ggplot(aes(x = PM2.5_ref, y = PM2.5_pred)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black", se = TRUE, fullrange = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(method ~ reference) +
  xlim(0, 20) +
  ylim(0, 20) +
  labs(
    title = "SPS30 vs Reference PM2.5 Concentrations",
    subtitle = "Calibrated values from the SPS30 sensor",
    x = "Reference PM2.5 (µg/m³)",
    y = "SPS30 PM2.5 (µg/m³)"
  ) +
  theme_minimal()

ggsave("plots/supplementary/scatter_clean_air.png",
  plot = p_clean_air,
  width = 8, height = 6, dpi = 600, bg = "white"
)

p_clean_air
```

```{r, prepare table 1s}
# inspect the number of observations per sensor per exposure source:
table1s_data_clean <- all_data |>
  count(sensor, source) |>
  mutate(
    source = factor(source),
    source = fct_recode(
      source,
      "Candles (high RH)" = "candles_high_RH",
      "Cooking (high RH)" = "cooking_high_RH",
      "Candles" = "candles",
      "Cooking" = "cooking",
      "Cigarette" = "cigarette",
      "Clean air" = "clean_air"
    )
  ) |>
  group_by(sensor, source) |>
  pivot_wider(names_from = source, values_from = n, values_fill = 0) |>
  rowwise() |>
  mutate(
    Sensor = as.character(sensor),
    Total = sum(c_across(where(is.numeric) | where(is.integer)))
  )
```

```{r, table 1 output}
# Build grouped GT table
table1s_gt <- table1s_data_clean |>
  gt() |>
  tab_header(
    title = "Observations per sensor per source"
  )

# To view in RStudio
table1s_gt

# To save as Word/HTML/PDF if needed
gtsave(table1s_gt, "tables/supplementary/table1s.docx", overwrite = TRUE) # Word
```



```{r, table 3s supplementary output}
table4s_data <- by_source_data |>
  filter(PM2.5_ref >= 5) |>
  mutate(
    PM2.5_ref = ifelse(PM2.5_ref == 0, 0.001, PM2.5_ref) # avoid division by zero
  ) |>
  group_by(reference, method, source) |>
  summarise(
    # core error terms
    MAE = mean(abs(error), na.rm = TRUE),
    RMSE = sqrt(mean(error^2, na.rm = TRUE)),
    MeanError = mean(error, na.rm = TRUE),
    sd_error = sd(error, na.rm = TRUE),

    # relative measures
    mean_ref = mean(PM2.5_ref, na.rm = TRUE),
    mean_sps30 = mean(PM2.5_pred, na.rm = TRUE),
    err_pct = mean(error / PM2.5_ref * 100, na.rm = TRUE),

    # correlation and variation
    r = cor(PM2.5_pred, PM2.5_ref, use = "complete.obs"),
    r2 = r^2,
    cv = sd(error, na.rm = TRUE) / mean(error, na.rm = TRUE),

    # limits of agreement
    LoA_Low = MeanError - 1.96 * sd_error,
    LoA_High = MeanError + 1.96 * sd_error,
    .groups = "drop"
  )


# Clean up labels and rounding
table4S_data_clean <- table4s_data |>
  mutate(
    r = round(r, 2),
    MAE = round(MAE, 1),
    RMSE = round(RMSE, 1),
    cv = round(cv, 2),
    MeanError = round(MeanError, 1),
    LoA_Low = round(LoA_Low, 1),
    LoA_High = round(LoA_High, 1)
  ) |>
  select(
    Source = source,
    Reference = reference,
    Method = method,
    r,
    MAE,
    RMSE,
    err_pct,
    `Mean Error` = MeanError,
    `LoA Low` = LoA_Low,
    `LoA High` = LoA_High
  ) |>
  arrange(Source, Reference, Method)

# Build grouped GT table
table4S_gt <- table4S_data_clean |>
  gt(groupname_col = "Source") |>
  tab_header(
    title = "Performance metrics for SPS30 sensor PM₂.₅ concentrations",
    subtitle = "Against reference instruments (SMPS + APS, DustTrak), stratified by source type and calibration method"
  ) |>
  cols_label(
    Reference = "Reference",
    Method = "Method",
    r = "Pearson´s r",
    MAE = "MAE (µg/m³)",
    RMSE = "RMSE (µg/m³)",
    err_pct = "Mean Relative Error (%)",
    `Mean Error` = "Mean Error (µg/m³)",
    `LoA Low` = "LoA Low (µg/m³)",
    `LoA High` = "LoA High (µg/m³)"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  )


# To view in RStudio
table4S_gt

# To save as Word/HTML/PDF if needed
gtsave(table4S_gt, "tables/supplementary/table4S.docx", overwrite = TRUE) # Word
```
# Generate Table 3S summarizing mean and sd from Bland Altman plot with loso by sensor data

```{r, table 3S output}
table3s_data <- by_sensor_data |>
  filter(PM2.5_ref >= 5) |>
  mutate(
    PM2.5_ref = ifelse(PM2.5_ref == 0, 0.001, PM2.5_ref) # avoid division by zero
  ) |>
  group_by(reference, method, sensor) |>
  summarise(
    # core error terms
    MAE = mean(abs(error), na.rm = TRUE),
    RMSE = sqrt(mean(error^2, na.rm = TRUE)),
    MeanError = mean(error, na.rm = TRUE),
    sd_error = sd(error, na.rm = TRUE),

    # relative measures
    mean_ref = mean(PM2.5_ref, na.rm = TRUE),
    mean_sps30 = mean(PM2.5_pred, na.rm = TRUE),
    err_pct = mean(error / PM2.5_ref * 100, na.rm = TRUE),

    # correlation and variation
    r = cor(PM2.5_pred, PM2.5_ref, use = "complete.obs"),
    r2 = r^2,
    cv = sd(error, na.rm = TRUE) / mean(error, na.rm = TRUE),

    # limits of agreement
    LoA_Low = MeanError - 1.96 * sd_error,
    LoA_High = MeanError + 1.96 * sd_error,
    .groups = "drop"
  )

# Clean up labels and rounding
table3S_data_clean <- table3s_data |>
  mutate(
    r = round(r, 2),
    MAE = round(MAE, 1),
    RMSE = round(RMSE, 1),
    cv = round(cv, 2),
    MeanError = round(MeanError, 1),
    LoA_Low = round(LoA_Low, 1),
    LoA_High = round(LoA_High, 1)
  ) |>
  select(
    Sensor = sensor,
    Reference = reference,
    Method = method,
    r, MAE, RMSE, err_pct,
    `Mean Error` = MeanError,
    `LoA Low` = LoA_Low,
    `LoA High` = LoA_High
  ) |>
  arrange(Sensor, Reference, Method)

# Build grouped GT table
table3S_gt <- table3S_data_clean |>
  gt(groupname_col = "Sensor") |>
  tab_header(
    title = "Performance metrics for SPS30 sensor PM₂.₅ concentrations",
    subtitle = "Against reference instruments (SMPS + APS, DustTrak), stratified by sensor node and calibration method"
  ) |>
  cols_label(
    Reference = "Reference",
    Method = "Method",
    r = "r",
    MAE = "MAE (µg/m³)",
    # cv = "Coefficient of variation",
    err_pct = "Mean Relative Error (%)",
    `Mean Error` = "Mean Error (µg/m³)",
    `LoA Low` = "LoA Low (µg/m³)",
    `LoA High` = "LoA High (µg/m³)"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  )


# To view in RStudio
table3S_gt

# To save as Word/HTML/PDF if needed
gtsave(table3S_gt, "tables/supplementary/table3S.docx", overwrite = TRUE) # Word
```
